Supporting Information : The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2019-05-02 SupportingInformation.Rmd
various edits, getting ready for submission

---
title: "Supporting Information : The importance of a complete statistical description of dependence between variables for ecology and related fields"
author: "Shyamolina Ghosh, Lawrence W. Sheppard, Mark T. Holder, Terrance E. Loecke, Philip C. Reid, Daniel C. Reuman"
geometry: "left=1cm,right=1cm,top=2.2cm,bottom=2.2cm"

@@

<!--***DAN: This is a dumping ground at the moment, but needs to be organized into a section that gives all the necessary details about the data-->

<!--additional details about Cedar creek data, etc.-->
Above-ground biomass and percent cover data were obtained from long-term (? or short-term)
biodiversity experiment (E120) conducted in Cedar Creek Ecosystem Science Reserve by Tilman *et al*.
In 1994, 168 plots (each 9 m $\times$ 9 m) of 39, 35, 29, 30 and 35 replicates,
were seeded with 1, 2, 4, 8, or 16 perennials, respectively, chosen randomly from a pool of 
18 species (4 species, each, of C4 grasses, C3 grasses, legumes, non-legume forbs; 2 species of woody
plants). All plots received, in total, 10 $gm^{-2}$ of pure live seed in May 1994 and 5 $gm^{-2}$ in May, 1995 
with seed mass divided equally among species. Treatments were maintained by weeding manually 3-4 times a year 
and sometimes selective herbicides were used to set up the monocultures either with only grasses or with only forbs.
We downloaded the data on July, 2018 from Cedar Creek website \url{http://www.cedarcreek.umn.edu/research/experiments/e120} and 
analyzed those [data](https://github.com/sghosh89/BIVAN/tree/master/Data/Ceder_creek_e120) for 6 common years 
(1996-2000, 2007) when both biomass and percent cover data were available. Each plots were annually sampled by clipping 
different locations each year. For each available year, we computed above-ground biomass ($gm^{-2}$) 
and Shannon's biodiversity index (using percent cover data) averaged over the sampled strips/ quadrats taken from each 
of 168 plots.

<!--we need to add aphid and plankton data details, Lawrence will give those-->

<!--The CPR is a device which is towed across the ocean behind a ship, filtering plankton from seawater with a continuously
unwinding silk ribbon. The ribbon was cut into 4 inch segments, each corresponding to a particular time and
location in the ocean. Organisms were identified and counted (possibly with sub-sampling) microscopically [@Colebrook1979]. All the samples available for a given variable in a given area in a given month of a given year were
averaged, producing a monthly time series for each year. Where data for a given month was unavailable the missing value was replaced with a median value for that month (drawn from all years in which data was available).-->

 [@Sheppard2017] converted monthly values to yearly values by averaging over all months. But here we only allowed maximum 4 months in a year with missing data which were replaced by median value for that month (drawn from all years in which data was available). We used data from 14 locations which had at least 45 years time series data (see Figure \ref{fig_plankton_map}).

 The Great Miami Wetland Mitigation Bank  was at $36^{\circ}46^{'}51^{''}$ N, $84^{\circ}20^{'}26^{''}$ W) in Trotwood, Montgomery County, Ohio. Additional information on the history of the site and soil types is detailed in @Jarecke2016. Briefly, the closed static flux chamber method was used [@Holland1999] to measure $CH_{4}$ fluxes at 28 locations across a hydrologic gradient (upland to seasonal wetland) at daily to weekly interval from September 2015 to September 2016. **See and** @smyth2019using, **pending for
additional details???** on greenhouse gas flux methods for this site. We considered 13 locations that had at least 50 dates data.

@@

Once character evolution was simulated 100 times for each dependence structure, 
we had 817 bivariate characters for each of 500 simulations. We thus analyzed 500 corresponding 
bivariate copulas (one for each simulation), each comprised of 817 points. 
To assess the symmetry or asymmetry of tail dependence for each copula, we computed 
the three non-parametric statistics $\cor_{0,0.2}-\cor{0.8,1}$, $\Ps_{0,0.2}-\Ps{0.8,1}$,
and $\Dsq_{0.8,1}-\Dsq_{0,0.2}$ for each copula. For each statistic and for each
dependence structure, we made a histogram (Fig. \ref{fig_cop_evo}). Values of these statistics less than 0
indicate lower-tail dependence in characters of extant species, whereas values greater
than 0 indicate upper-tail dependence. Asymmetry of tail dependence in evolutionary changes to characters
was associated with similar asymmetry of tail dependence in extant characters.

@@

@@

<!--Following chunk is to examine the hypothesis for third probable mechanism : 
asymmetric tail dependence in evolutionary changes in bivariate characters can give
rise to similarly asymmetric tail dependence in character distributions across phylogeny tips-->

```{r mechanism_evo_cop, echo=F, results="hide", cache=T, cache.extra=list(seed,mtime("explore_evo_cop.R"),mtime("CopulaFunctions_flexible.R"))}
set.seed(seed)

@@

<!--adding plots showing results to confirm hypothesis behind third mechanism-->
\begin{figure}[!h]
\textbf{ \hspace{6.5 cm} (A) \hspace{2 cm} (B) \hspace{2 cm} (C)} \\
\vspace{-0.5 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/ExtremeL_hist_taildep.pdf}\\
\end{center}
\textbf{ \hspace{6.5 cm} (D) \hspace{2 cm} (E) \hspace{2 cm} (F)} \\
\vspace{-0.5 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/SomewhatL_hist_taildep.pdf}\\
\end{center}
\textbf{ \hspace{6.5 cm} (G) \hspace{2 cm} (H) \hspace{2 cm} (I)} \\
\vspace{-0.7 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/BivNormal_hist_taildep.pdf}\\
\end{center}
\textbf{ \hspace{6.5 cm} (J) \hspace{2 cm} (K) \hspace{2 cm} (L)} \\
\vspace{-0.5 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/SomewhatR_hist_taildep.pdf}\\
\end{center}
\textbf{ \hspace{6.5 cm} (M) \hspace{2 cm} (N) \hspace{2 cm} (O)} \\
\vspace{-0.5 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/ExtremeR_hist_taildep.pdf}\\
\caption[Non-parametric statistics on a mechanism of tail dependence in character distributions]{Three 
measures of asymmetry of tail dependence ($\cor_{0,0.2}-\cor_{0.8,1}$ (A, D, G, J, M), 
$\Ps_{0,0.2}-\Ps_{0.8,1}$ (B, E, H, K, N), and $\Dsq_{0.8,1}-\Dsq_{0,0.2}$ (C, F, I, L, O)) between two characters, across extant species,  were 
computed for each of 100 simulations of mammalian evolution for each of five types of tail dependence between 
evolutionary changes in the characters (extreme left-tail dependence (A-C), moderate left-tail dependence (D-F), 
symmetry of tail dependence (G-I), moderate right-tail dependence (J-L), and extreme right-tail dependence 
(M-O)). Asymmetry of tail dependence in evolutionary changes in characters tended to beget similar
asymmetry in the characters across extant species. For additional details see 
Appendix \ref{evol_mech}. The number in the upper right of each panel indicates how many statistics out of $100$ were less than or equal to $0$. Values substantially lower (respectively, higher) than $50$ therefore indicate greater left (respectively, right) tail-dependence between characters.\label{fig_cop_evo}}
\end{center}
\end{figure}

@@

>>>>

---
title: "Supporting Information : The importance of a complete statistical description of dependence between variables for ecology and related fields"
author: "Shyamolina Ghosh, Lawrence W. Sheppard, Mark T. Holder, Terrance E. Loecke, Philip C. Reid, James D. Bever, Daniel C. Reuman"
geometry: "left=1cm,right=1cm,top=2.2cm,bottom=2.2cm"

@@

\textbf{Population data, plankton abundances.} 
The CPR plankton data presented in @Sheppard2017 was aggregated in the following way. Fifty 
$2^\circ$ by $2^\circ$ location boxes were examined between $-10^\circ$ and $10^\circ$ longitude 
and between $50^\circ$ and $60^\circ$ latitude (Fig. \ref{fig_plankton_map}). For each species, for each box, 
for each year, for each month, the average of any available CPR samples was taken to give monthly 
time series of population densities with some missing values. Missing monthly values were then 
replaced with the median value for that species for that location for that month. The data for each 
species for each box for each year was then averaged over all twelve months, and 26 usable annual 
time series were identified for wavelet analysis. The CPR plankton data used here is aggregated in 
the same way, except that for any species, box and year for which more than 4 months data were missing, 
missing values were not replaced with medians and those species, box, year combinations were not used.
We used only data from the 14 locations which had at least 45 years of annual data subject to this 
restriction, shown by blue markers in Fig. \ref{fig_plankton_map}. The CPR data are described in
great detail elsewhere, e.g., by  @Batten2003 and @Raitsos2014.

\textbf{Community-level data, Cedar Creek.} Above-ground biomass and percent cover data were 
obtained from biodiversity experiment E120 conducted at the Cedar Creek Ecosystem Science Reserve 
by Tilman *et al*. In 1994, 168 plots (each 9 m $\times$ 9 m) of 39, 35, 29, 30 and 35 replicates,
were seeded with 1, 2, 4, 8, or 16 perennials, respectively, chosen randomly from a pool of 
18 species (4 species, each, of C4 grasses, C3 grasses, legumes, non-legume forbs; and 2 species of woody
plants). All plots received, in total, 10 $gm^{-2}$ of pure live seed in May 1994 and 5 $gm^{-2}$ in May, 1995 
with seed mass divided equally among species. Treatments were maintained by weeding manually 3-4 times a year 
and sometimes selective herbicides were used. We downloaded the data on July, 2018 from the Cedar Creek website \url{http://www.cedarcreek.umn.edu/research/experiments/e120} and 
analyzed those data for 6 years 
(1996-2000, 2007) for which both biomass and percent cover data were available. Plots were annually 
sampled by clipping 
different locations each year. For each available year, we analyzed above-ground biomass ($gm^{-2}$) 
and Shannon's biodiversity index (computed using percent cover data) averaged over the 
sampled strips/quadrats taken from each 
of 168 plots.

\textbf{Ecosystem functioning data on methane flux.} The Great Miami Wetland Mitigation Bank  was at 
$36^{\circ}46^{'}51^{''}$ N, $84^{\circ}20^{'}26^{''}$ W, in Trotwood, Montgomery County, Ohio. Additional information 
on the history of the site and soil types is detailed in @Jarecke2016. Briefly, the closed static flux chamber method 
was used [@Holland1999] to measure $\text{CH}_{4}$ fluxes at 28 locations across a hydrologic gradient (upland to 
seasonal wetland) at daily to weekly interval from September 2015 to September 2016. See @smyth2019using for
additional details on greenhouse gas flux methods for this site. We considered 13 locations for which data were
gathered for at least 50 dates.

@@

Once character evolution was simulated 100 times for each dependence structure, 
we had 817 bivariate characters for each of 500 simulations. We thus analyzed 500 corresponding 
bivariate copulas (one for each simulation), each comprised of 817 points. 
To assess the symmetry or asymmetry of tail dependence for each copula, we computed 
the three non-parametric statistics $\cor_{0,0.2}-\cor{0.8,1}$, $\Ps_{0,0.2}-\Ps{0.8,1}$,
and $\Dsq_{0.8,1}-\Dsq_{0,0.2}$ for each copula. For each statistic and for each
dependence structure, we made a histogram of values 
(Fig. \ref{fig_cop_evo}). Asymmetry of tail dependence in evolutionary changes to characters
was associated with similar asymmetry of tail dependence in extant characters.

@@

Additionally, character evolution was simulated using one random draw from the relevant matrix
per phylogeny branch. This was principally because branch lengths are often hugely
uncertain. If branch lengths were well know, alternative simulation strategies may
include selecting one draw from the matrix of evolutionary changes per unit time, 
or selecting one draw for the whole branch but rescaling variance according to the 
length of the branch. These choices amount to the same thing for normal copula
structure, but not otherwise. Modelling choices such as these may have influenced
our results. Additional research also seems warranted testing the realism 
of our hypothesized mechanism and simulations.

@@

<!--Following chunk is to examine the hypothesis for third probable mechanism : 
asymmetric tail dependence in evolutionary changes in bivariate characters can give
rise to similarly asymmetric tail dependence in character distributions across phylogeny tips-->
```{r mechanism_evo_cop, echo=F, results="hide", cache=T, cache.extra=list(seed,mtime("explore_evo_cop.R"),mtime("CopulaFunctions_flexible.R"))}
set.seed(seed)

@@

<!--adding plots showing results to confirm hypothesis behind third mechanism-->
\begin{figure}[!h]
\textbf{ \hspace{6.5 cm} (A) \hspace{2 cm} (B) \hspace{2 cm} (C)} \\
\vspace{-0.5 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/ExtremeL_hist_taildep.pdf}\\
\end{center}
\textbf{ \hspace{6.5 cm} (D) \hspace{2 cm} (E) \hspace{2 cm} (F)} \\
\vspace{-0.5 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/SomewhatL_hist_taildep.pdf}\\
\end{center}
\textbf{ \hspace{6.5 cm} (G) \hspace{2 cm} (H) \hspace{2 cm} (I)} \\
\vspace{-0.7 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/BivNormal_hist_taildep.pdf}\\
\end{center}
\textbf{ \hspace{6.5 cm} (J) \hspace{2 cm} (K) \hspace{2 cm} (L)} \\
\vspace{-0.5 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/SomewhatR_hist_taildep.pdf}\\
\end{center}
\textbf{ \hspace{6.5 cm} (M) \hspace{2 cm} (N) \hspace{2 cm} (O)} \\
\vspace{-0.5 cm}
\begin{center}
\includegraphics[width=9 cm]{./Results/BMR_results/evo_cop/ExtremeR_hist_taildep.pdf}\\
\caption[Non-parametric statistics on a mechanism of tail dependence in character distributions]{Three 
measures of asymmetry of tail dependence ($\cor_{0,0.2}-\cor_{0.8,1}$ (A, D, G, J, M), 
$\Ps_{0,0.2}-\Ps_{0.8,1}$ (B, E, H, K, N), and $\Dsq_{0.8,1}-\Dsq_{0,0.2}$ (C, F, I, L, O)) between two characters, across extant species,  were 
computed for each of 100 simulations of mammalian character evolution for each of five types of tail dependence between 
evolutionary changes in the characters (extreme left-tail dependence (A-C), moderate left-tail dependence (D-F), 
symmetry of tail dependence (G-I), moderate right-tail dependence (J-L), and extreme right-tail dependence 
(M-O)). Asymmetry of tail dependence in evolutionary changes in characters tended to beget similar
asymmetry in the characters across extant species. For additional details see 
Appendix \ref{evol_mech}. The number in the upper right of each panel indicates how many statistics out of $100$ were less than or equal to $0$. Values substantially lower (respectively, higher) than $50$ therefore indicate greater left (respectively, right) tail-dependence between characters.\label{fig_cop_evo}}
\end{center}
\end{figure}

@@

# A means by which missing data can influence perceived copula structure\label{sect:missingdata}

```{r missing_data, echo=FALSE, results="hide", cache=TRUE}
set.seed(101)

#generate data from a normal copula
ncopparm<-0.8
ncop<-normalCopula(ncopparm,2)
d<-rCopula(5000,ncop)

#censor it - randomly delete many of the points that are in the upper right 
#triangle of the copula plot
inds<-which(d[,2]>-d[,1]+1)
fracdelete<-4/5
inds<-inds[sample(1:length(inds),size=round(fracdelete*length(inds)),replace=FALSE)]
dc<-d[-inds,]

#reperform the ranking
dc<-pobs(dc)

#now make a plot
panwd<-2
panht<-panwd
xaxht<-.6
yaxwd<-xaxht
gap<-0.1
totwd<-yaxwd+2*panwd+2*gap
totht<-xaxht+panht+gap
pdf(file="./Results/MissingDataInfluencesCopula.pdf",width=totwd,height=totht)
par(fig=c(yaxwd/totwd, #panel 1 - the original data
          (yaxwd+panwd)/totwd,
          (xaxht)/totht,
          (xaxht+panht)/totht),
    mai=c(0,0,0,0),mgp=c(3,.15,0),tcl=-.25)
plot(d[,1],d[,2],pch=20,cex=.5,xlim=c(0,1),ylim=c(0,1))
text(0,1,"A",adj=c(0,1),cex=1.5)
mtext("x",1,1.5,cex=2)
mtext("y",2,1.5,cex=2)
par(fig=c((yaxwd+panwd+gap)/totwd, #panel 2 - the censored data, re-ranked
          (yaxwd+2*panwd+gap)/totwd,
          (xaxht)/totht,
          (xaxht+panht)/totht),
    mai=c(0,0,0,0),mgp=c(3,.15,0),tcl=-.25,new=TRUE)
plot(dc[,1],dc[,2],pch=20,cex=.5,xlim=c(0,1),ylim=c(0,1),yaxt="n")
axis(2,labels=FALSE)
text(0,1,"B",adj=c(0,1),cex=1.5)
mtext("x",1,1.5,cex=2)
dev.off()
```

To illustrate the potential for missing data to influence perceived
copula structure, we first generated 5000 points from a bivariate N copula
with parameter `r ncopparm` (Fig. \ref{fig:missingdata}A), and then 
randomly deleted `r 100*fracdelete`% of the points that occurred in the
upper-right triangle of the copula plot. Data were then re-ranked as
described in the Introduction to produce a new copula plot based
only on the censored dataset (Fig. \ref{fig:missingdata}B). Copula
structure in the tails was affected. Although real datasets have 
typically not been censored in the precise way we censored our
artificial data, it may be common for species characters to be 
measured on a subset of the extant species from the clade of interest,
and the probability of a species having been included in the dataset
may depend on one or both of the character values. 
For instance, if larger or smaller bird or mammal species were more likely
to have been measured for body mass or BMR, such accidental 
censoring of the data could have influenced our measurements of tail 
dependence. We have no reason to believe this is the case for our 
BMR versus body mass datasets, but we admit the possibility. Any systematic 
use of copulas to study life-history tradeoffs should take the possibility
into account of artefacts such as described here.

\begin{figure}
\begin{center}
\includegraphics[width=7in]{./Results/MissingDataInfluencesCopula.pdf}
\caption[Missing data can influence perceived copula structure]{Missing
data can influence perceived copula structure. Points were randomly 
removed from the upper right portion of the copula dataset pictured in A, and
then remaining values were re-ranked to make B. Thus removing data in a 
non-random fashion can influence copula structure and tail dependence. See 
Appendix \ref{sect:missingdata} for details. \label{fig:missingdata}}
\end{center}
\end{figure}
